Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
6.1.1 WIP

\newcommand{\NUMFUNCTIONS}{20,214\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
\newcommand{\TOTNUMSIGS}{21,730\xspace}

@@

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  binom.profile <- function(x, n, ...) {
   xn <- cbind(x = x, n = n)
   ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
   x <- xn[ok, "x"]
   n <- xn[ok, "n"]
\end{lstlisting}

The \code{binom} package is a popular R package for computing confidence
intervals, and the \code{binom.profile} function does so using the profile
likelihood.  This code snippet highlights a data sanitization pattern: the
programmer first binds the vectors into a matrix, with one column for each
of \code{x} and \code{n}, then finds rows where both columns are not \NA,
then extracts non-\NA values and stores them into \code{x} and \code{n}
respectively.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  hankel.matrix <- function( n, x ) { 
  ### n = a positive integer value for the order of the Hankel matrix
  ### x = an order 2 * n + 1 vector of numeric values
    ...
    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
    m <- length( x )
    if ( m < n )  stop( "length of argument x is less than n" )
\end{lstlisting}

The \code{hankel.matrix} function takes two arguments.  The function has a
number of interesting checks: \code{n} undergoes an integer type check;
\code{x} undergoes a vector type check; and \code{n} is expected to be a
scalar, as if it were a vector, the code \code{m < n} would essentially
check \code{m < n[i]} for each element in \code{n}, returning a vector of
comparison results, and the conditional of an if must be a scalar else R
generates a warning.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
                             na.rm = FALSE, ...) {
  if (!is.matrix(x)) 
    .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
\end{lstlisting}

The \code{rowWeightedMeans} function calculates the weighted means of rows
or columns of a matrix \code{x}.  This function exhibits good practice in
type checking arguments, rightly producing a message to the user indicating
that an explicit matrix should be passed.

To capture cases like this, we include a \code{class<matrix>} type in our
type annotation language.  If \code{rowWeightedMeans} were outfitted with
such a type, \contractr can perform the check present in this code
automatically, notifying the users passing ill-typed arguments.

Thus far we've seen instances where we introduce types that subsume existing
argument validation code, but our intention is not to subsume {\it all} such
code.

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
  rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
    n <- ncol(x)
    if (length(w) != n) 
       stop("The length of 'w' does not match columns in 'x': ")
\end{lstlisting}

@@

We observed the use of consistently-named lists in \AT{a number of
  functions}.  In R, ``fields'' of named lists can be accessed using the
\code{$} operator.

>>>>

\newcommand{\NUMFUNCTIONS}{20,214\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
\newcommand{\TOTNUMSIGS}{21,730\xspace}
\newcommand{\PROPSTRUCTSDYN}{44.46\%\xspace}

@@

Now, we aim to enhance the quantitative analysis of our language design by looking at select code snippets from the corpus we analyzed.
We will highlight some of the design decisions that we made, give code examples, and accompany them with function signatures generated by \typetracer.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <x: dbl[], n: dbl[], conf.level: dbl, maxsteps: dbl,
#'         del: dbl, bayes: lgl, plot: lgl, ...> => class<data.frame>
binom.profile <- function(x, n, conf.level = 0.95, maxsteps = 50,
                            del = zmax / 5, bayes = TRUE, plot = FALSE, ...) {
   xn <- cbind(x = x, n = n)
   ok <- !is.na(xn[, 1]) & !is.na(xn[, 2])
   x <- xn[ok, "x"]
   n <- xn[ok, "n"]
\end{lstlisting}

The \code{binom} package is a popular R package for computing confidence
intervals, and the \code{binom.profile} function does so using the profile
likelihood.  This code snippet highlights a data sanitization pattern: the
programmer first binds the vectors into a matrix, with one column for each
of \code{x} and \code{n}, then finds rows where both columns are not \NA,
then extracts non-\NA values and stores them into \code{x} and \code{n}
respectively.We see from the type that arguments \code{x} and \code{n} are intended to be free of \NA values.
\contractr will check them for compliance, and warn programmers of function misuse.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <n: dbl, x: int[]> => class<matrix>
hankel.matrix <- function( n, x ) { 
  ### n = a positive integer value for the order of the Hankel matrix
  ### x = an order 2 * n + 1 vector of numeric values
    ...
    if ( n != trunc( n ) )  stop( "argument n ix not an integer" )
    if ( !is.vector( x ) )  stop( "argument x is not a vector" )
    m <- length( x )
    if ( m < n )  stop( "length of argument x is less than n" )
\end{lstlisting}

The \code{hankel.matrix} function takes two arguments.  The function has a
number of interesting checks: \code{n} undergoes an integer type check;
\code{x} undergoes a vector type check; and \code{n} is expected to be a
scalar, as if it were a vector, the code \code{m < n} would essentially
check \code{m < n[i]} for each element in \code{n}, returning a vector of
comparison results, and the conditional of an if must be a scalar else R
generates a warning. The type for this function indicates that \code{x} must be
a vector, though here the type for \code{n} is found to be \dbl---this is likely due
to numeric values defaulting to \dbl type in R, as e.g. \code{typeof(2)} is \dbl.

@@

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
#' @type <x: class<matrix>, w: ? ^dbl[], rows: null, cols: null, na.rm: lgl, ...> => ^dbl[]
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
                             na.rm = FALSE, ...) {
  if (!is.matrix(x)) 
    .Defunct(msg = sprintf("Argument 'x' should be a matrix.)
\end{lstlisting}

The \code{rowWeightedMeans} function calculates the weighted means of rows
or columns of a matrix \code{x}.  This function exhibits good practice in
type checking arguments, rightly producing a message to the user indicating
that an explicit matrix should be passed. We captured this with our class types, 
and indeed \typetracer inferred a good type for this function, allowing \contractr to 
perform the check present in this code automatically, notifying the users passing ill-typed arguments.

Thus far we've seen instances where we introduce types that subsume existing
argument validation code, but our intention is not to subsume {\it all} such
code. Consider:

\begin{lstlisting}[basicstyle=\footnotesize\ttfamily]
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL, na.rm = FALSE) {
    ...
    n <- ncol(x)
    if (length(w) != n) 
       stop("The length of 'w' does not match columns in 'x': ")
\end{lstlisting}

@@

\typetracer observed that \PROPSTRUCTSDYN of lists had a \code{names} attribute, and in R, ``fields'' of named lists can be accessed using the
\code{$} operator.
